---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pangoRo

<!-- badges: start -->

[![R-CMD-check](https://github.com/al-obrien/pangoRo/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/al-obrien/pangoRo/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

COVID-19 lineage names can be confusing to navigate; there are many aliases and if you want to catch them all to examine further, it helps to have some additional tools...

{pangoRo} is an R package to support interacting with [PANGO lineage](https://cov-lineages.org/index.html) information. The core functionality was inspired by a similar package called [pango_aliaser](https://github.com/corneliusroemer/pango_aliasor) created by Cornelius Roemer for the Python language.

## Installation

You can install {pangoRo} from GitHub:

``` r
remotes::install_github('al-obrien/pangoRo')
```

## Examples

The basic usage of {pangoRo} is to expand, collapse, and sort COVID-19 lineages. Start by creating the *pangoro* object that links to the latest (or cached) PANGO reference. This is then passed to subsequent operations as reference.

```{r example}
library(pangoRo)

# Create pangoro object
my_pangoro <- pangoro()
```

### Collapse

With a vector of PANGO lineages, provide fully collapsed output.

```{r}
# Vector of COVID-19 lineages to collapse
cov_lin <- c('B.1.617.2', 'BL.2', 'B.1.1.529.2.75.1.2', 'BA.2.75.1.2', 'XD.1')

# Collapse lineage names as far as possible
collapse_pangoro(my_pangoro, cov_lin)
```

Can also define how far to collapse each input.

```{r}
collapse_pangoro(my_pangoro, cov_lin, max_level = 1)
```


### Expand

```{r}
# Vector of COVID-19 lineages to expand
cov_lin <- c('B.1.617.2', 'B.1.617.2.6', 'AY.4', 'AY.39', 'BL.2', 'BA.1', 'AY.2', 'XD.1')

# Expand lineage names as far as possible
exp_lin <- expand_pangoro(my_pangoro, cov_lin)
exp_lin
```

### Sort

Perform a pseudo-sort on the lineage names.

```{r}
# Sort lineages
sort_pangoro(my_pangoro, exp_lin)
```

Split the lineages by their lowest alias codes and sort within each grouping

```{r}
collapsed_full <- collapse_pangoro(my_pangoro, cov_lin, aliase_parent = TRUE) 
grps <-  split(collapsed_full, sapply(strsplit(collapsed_full, split = '\\.'), `[[`, 1))
lapply(grps, function(x) sort_pangoro(my_pangoro, x))
```

